The quench action approach for integrable models: A Monte Carlo study

### 2015-10-25 draft/fs_qa_v4.tex
rev 25/10/2015

\title{The quench action approach for integrable models: A Monte Carlo implementation}

@@

Understanding the out-of-equilibrium dynamics in {\it isolated} quantum 
many-body systems is one of the most intriguing research topics in 
contemporary physics~\cite{polkovnikov-2011}, both theoretically and 
experimentally. The most common out-of-equilibrium protocol is the {\it 
quantum quench} in which a system is initially prepared in an eigenstate 
$|\Psi_0\rangle$ of a many-body hamiltonian ${\mathcal H}$. Then a global 
parameter of the hamiltonian is suddenly changed, and the system is let 
to evolve unitarily under a new Hamiltonian ${\mathcal H}'$.

While there is now compelling evidence that at long times after the 
quench a steady state arises, its nature is still highly debated. 
In the thermodynamic limit it is natural to expect that the steady-state 
value of any local observable ${\mathcal O}$ is given by the so-called 
diagonal ensemble as 
%
\begin{equation}
\label{d-ensemble}
\langle{\mathcal O}\rangle_{DE}=\sum\limits_{\lambda}|\langle\Psi_0|\lambda
\rangle|^2\langle\lambda|{\mathcal O}|\lambda\rangle,
\end{equation}
%
where the sum is over the eigenstates $|\lambda\rangle$ of the post-quench 
final hamiltonian ${\mathcal H}'$. For generic, i.e., non-integrable, models 
the Eigenstate Thermalization Hypothesis ensures that the ensemble~\eref{d-ensemble} 
becomes equivalent to the Gibbs (thermal) ensemble, implying 
%
\begin{equation}
\langle{\mathcal O}\rangle_{DE}=\langle{\mathcal O}\rangle_{Gibbs}\equiv
\frac{1}{Z}\Tr({\mathcal O}\rho^{Gibbs}),\quad\textrm{with}\quad\rho^{Gibbs}
=e^{-\beta{\mathcal H}'},
\end{equation}
%
where $\beta$ is the inverse temperature of the system and $Z$ a normalization 
factor. For one-dimensional integrable models, the presence of local or quasi-local 
integrals of motion strongly affects the dynamics, preventing the onset of 
thermal behavior at long times. It has been suggested that the long time 
steady state behavior is described by the so-called Generalized Gibbs 
ensemble as 
%
\begin{equation}
\langle{\mathcal O}\rangle_{GGE}=\frac{1}{Z}\Tr({\mathcal O}\rho^{GGE}), 
\quad\textrm{with}\quad\rho^{GGE}\equiv\frac{1}{Z}\exp\Big(-\sum_j\beta_jQ_j\Big)
\end{equation}
%
We show that it is possible to numerically simulate the Quench Action approach 
combining Monte Carlo methods and Bethe ansatz techniques.

@@

It is useful to rewrite~\eref{d-ensemble} as 
%
\begin{equation}
\label{qa-prob}
\langle{\mathcal O}\rangle=\sum\limits_{\alpha}\rho^{DE}(\alpha)
\langle\alpha|{\mathcal O}|\alpha\rangle,\quad\textrm{with}\quad 
\rho^{DE}(\alpha)=\exp(-2\Re{\mathcal E}(\alpha)).
\end{equation}
%
Here $\rho^{DE}(\alpha)$ is a diagonal ensemble density matrix, and ${\mathcal E}(\alpha)
\equiv-\log\langle\alpha|\Psi_0\rangle$, with $\Re$ denoting the real part.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overlap between the Bethe states and some initial states} 
\label{sec:2}

@@

Here we detail the Bethe ansatz results for the overlap of the Bethe states (cf.~\eref{ba-eig}) 
with the zero-momentum (one-site shift invariant) N\'eel state $|N\rangle$ (cf.~\eref{neel}) 
and the Majumdar-Ghosh (MG) $|MG\rangle$ state (cf.~\eref{mg}). We stard discussing the N\'eel 
state. Due to the zero-momentum constraint, only parity-invariant Bethe states can have non-zero 
N\'eel overlap~\cite{brockmann-2014,wouters-2014}. The corresponding solutions of the Bethe 
equations~\eref{ba-eq} contain only pairs of rapidities with opposite sign.  Here we denote the 
generic parity-invariant rapidity configuration as $|\{\pm\tilde\lambda_j\}_{j=1}^m,n_\infty
\rangle$, i.e., considering only positive rapidities (as stressed by the tilde in $\tilde
\lambda_j$). Here $m$ is the number of rapidity pairs. Since the N\'eel state is not invariant 
under $SU(2)$ rotations, eigenstates with infinite rapidities can have non-zero N\'eel overlaps.
We denote the number of infinite rapidities as $N_{\infty}$. Note that one has $M=L/2=
N_\infty+2m$. The density of infinite rapidities is denoted as $n_\infty\equiv N_\infty/L$. 
The overlap between the Bethe states and the Neel state $|N\rangle$ reads~\cite{brockmann-2014,
pozsgay-2014} 
%
\begin{equation}
\label{neel-ov}
\frac{\langle N|\{\pm\tilde\lambda_j\}_{j=1}^m,n_\infty\rangle}{|||\{\tilde\lambda_j\}_{j=1}^m,
n_\infty\rangle||}=\frac{\sqrt{2}N_{\infty}!}{\sqrt{(2N_\infty)!}}\left[\prod_{j=1}^m
\frac{\sqrt{\tilde\lambda_j^2+1}}{4\tilde\lambda_j}\right]\sqrt{\frac{\textrm{det}_m(G^+)}{
\textrm{det}_m(G^-)}}.
\end{equation}
%
The matrix $G^\pm$ is  defined as  
%
\begin{equation}
\label{G-pm}
\fl\quad G^{\pm}_{jk}=\delta_{jk}\Big(LK_{1/2}(\tilde\lambda_j)-\sum\limits_{l=1}^mK_1^+(
\tilde\lambda_j,\tilde\lambda_l)\Big)+K_{1}^{\pm}(\tilde\lambda_j,\tilde\lambda_k),
\quad\,j,k=1,\dots,m, 
\end{equation}
%
where 
%
\begin{equation}
\label{K}
K_1^\pm(\lambda,\mu)=K_1(\lambda-\mu)\pm K_1(\lambda+\mu) \quad\textrm{with}\quad 
K_\alpha(\lambda)\equiv\frac{8\alpha}{\lambda^2+4\alpha^2}. 
\end{equation}
%
Note that our definitions of $K_{\alpha}(\lambda)$ differs from the one in 
Ref.~\cite{brockmann-2014}, due to a factor $2$ in the definition of the 
rapidities (cf.~\eref{str-hyp}).

@@

@@

>>>>

\title{The quench action approach for integrable models: A Monte Carlo study}

@@

Understanding the out-of-equilibrium dynamics in {\it isolated} quantum 
many-body systems is one of the most intriguing research topics in 
contemporary physics, both theoretically~\cite{polkovnikov-2011} and 
experimentally~\cite{bloch-2008,greiner-2002,kinoshita-2006,
hofferberth-2007,trotzky-2012,gring-2012,cheneau-2012,
schneider-2012,kunhert-2013,langen-2013,meinert-2013,fukuhara-2013,
ronzheimer-2013,braun-2014,langen-2015}. The most popular out-of-equilibrium 
protocol is that of the {\it quantum quench} in which a system is initially 
prepared in an eigenstate $|\Psi_0\rangle$ of a many-body hamiltonian ${\mathcal H}_0$. 
Then a global parameter is suddenly changed, and the system is let to evolve 
unitarily under a new hamiltonian ${\mathcal H}$. There is now compelling 
evidence that at long times after the quench an equilibrium steady state 
arises (see for instance Ref.~\cite{kinoshita-2006}).

In the thermodynamic limit, due to dephasing, it is natural to expect that the 
equilibrium value of any local observable ${\mathcal O}$ is described by the 
so-called diagonal ensemble as 
%
\begin{equation}
\label{d-ensemble}
\langle{\mathcal O}\rangle_{DE}=\sum\limits_{\lambda}|\langle\Psi_0|\lambda
\rangle|^2\langle\lambda|{\mathcal O}|\lambda\rangle,
\end{equation}
%
where the sum is over the eigenstates $|\lambda\rangle$ of the post-quench 
hamiltonian ${\mathcal H}$. For generic, i.e., non-integrable, models 
the Eigenstate Thermalization Hypothesis~\cite{deutsch-1991,srednicki-1994} 
(ETH) implies that the diagonal ensemble~\eref{d-ensemble} becomes equivalent 
to the usual Gibbs (thermal) ensemble. 
%, implying 
%
%\begin{equation}
%\langle{\mathcal O}\rangle_{DE}=\langle{\mathcal O}\rangle_{Gibbs}\equiv
%\frac{1}{Z}\Tr({\mathcal O}\rho^{Gibbs}),\quad\textrm{with}\quad\rho^{Gibbs}
%=e^{-\beta{\mathcal H}},
%\end{equation}
%
%where $\beta$ is the inverse temperature of the system and $Z$ a normalization 
%factor. 
In integrable models, however, the presence of an extensive number of local or 
quasi-local integrals of motion strongly affects the dynamics, preventing the onset of 
thermal behavior. It has been suggested in Ref.~\cite{rigol-2007,rigol-2008} that   
in this situation the post-quench steady-state can be described by the so-called 
Generalized Gibbs Ensemble ($GGE$) as 
%
\begin{equation}
\label{gge}
\langle{\mathcal O}\rangle_{GGE}=\frac{1}{Z}\Tr({\mathcal O}\rho^{GGE}), 
\quad\textrm{with}\quad\rho^{GGE}\equiv\frac{1}{Z}\exp\Big(-\sum_j\beta_jQ_j\Big). 
\end{equation}
%
Here $Q_j$ are mutually commuting conserved quantities, i.e., $[{\mathcal H},Q_j]=0\,
\forall j$ and $[Q_j,Q_k]=0\,\forall j,k$, whereas $\beta_j$ are Lagrange multipliers 
to be fixed by imposing that  $\langle Q_j\rangle_{GGE}=\langle\Psi_0|Q_j|\Psi_0\rangle$. 
The validity of the GGE has been confirmed in non-interacting theories~\cite{calabrese-2011,
calabrese-2012,fagotti-2013,kcc14,kcc14a}, whereas in interacting ones the scenario is 
still not settled. A recent work~\cite{ilievski-2015a} suggests that the $GGE$ 
description of the steady state  is complete provided that the so-called quasi-local 
charges~\cite{prosen-2014,pereira-2014,ilievski-2015} are included in~\eref{gge}.

Valuable insights into this issue have been provided by the so-called quench-action 
approach~\cite{caux-2013}. For Bethe ansatz integrable models this allows to construct 
the diagonal ensemble average~\eref{d-ensemble} in the thermodynamic limit, provided 
that the overlaps $\langle\lambda|\Psi_0\rangle$ are known. The physical idea is that 
for large system sizes it is possible to approximate the sum over the eigenstates 
in~\eref{d-ensemble} using a saddle point argument. Specifically, one starts with 
rewriting~\eref{d-ensemble} as 
%
\begin{equation}
\label{de1}
\langle{\mathcal O}\rangle_{DE}= \sum\limits_{\lambda}\rho^{DE}\langle\lambda|
{\mathcal O}|\lambda\rangle,\quad\textrm{with}\quad\rho^{DE}=\exp(-2\Re{\mathcal 
E}(\lambda)), 
\end{equation}
%
where ${\mathcal E}(\lambda)\equiv-\log\langle\lambda|\Psi_0\rangle$, and $\Re$ 
denoting the real part. For typical initial states $|\Psi_0\rangle$
one has ${\mathcal E}\propto L$, with $L$ the system size, reflecting  a typical 
vanishing of the overlaps as $\langle\Psi_0|\lambda\rangle\propto e^{-c L}$. 
Similar to the standard Thermodynamic Bethe Ansatz (TBA)~\cite{taka-book}, the 
extensivity of 
${\mathcal E}$ suggests that in the thermodynamic limit~\eref{de1} is dominated 
by a saddle point. Surprisingly, for some Bethe ansatz solvable models and some 
simple initial states standard TBA techniques, 
complemented with analytical knowledge of the overlaps $\langle\Psi_0|\lambda
\rangle$, allow to determine {\it analytically} this saddle point, and for some  
local observables, the diagonal ensemble expectation values~\eref{de1}, in the 
thermodynamic limit. This has been done succesfully for instance in the Lieb-Liniger 
model for the quench from the Bose-Einstein condensate~\cite{de-nardis-2014} 
(BEC) and for the $XXZ$ spin chain for the quench from the N\'eel state~\cite{wouters-2014,
brockmann-2014}. Interestingly, recently it has been suggested that the quench action 
approach could allow to reconstruct the full relaxation dynamics to the steady 
state~\cite{de-nardis-2015a,de-nardis-2015}.

%Formally,~\eref{de1} is equivalent to the standard Gibbs 
%ensemble after replacing ${\mathcal H}\to{\mathcal E}$. A crucial difference is 
%that while typical hamiltonians are local, ensuring the extensivity of the energy, 
%this is not guaranteed for ${\mathcal E}$.

Here we provide a Monte Carlo implementation of the quench action approach in the 
finite-size spin-$1/2$ Heisenberg chain ($XXX$ chain). combining Monte Carlo methods and Bethe ansatz techniques.

@@

%It is useful to rewrite~\eref{d-ensemble} as 
%
%\begin{equation}
%\label{qa-prob}
%\langle{\mathcal O}\rangle=\sum\limits_{\alpha}\rho^{DE}(\alpha)
%\langle\alpha|{\mathcal O}|\alpha\rangle,\quad\textrm{with}\quad 
%\rho^{DE}(\alpha)=\exp(-2\Re{\mathcal E}(\alpha)).
%\end{equation}
%
%Here $\rho^{DE}(\alpha)$ is a diagonal ensemble density matrix, and ${\mathcal E}(\alpha)
%\equiv-\log\langle\alpha|\Psi_0\rangle$, with $\Re$ denoting the real part.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overlap between the Bethe states and some simple roduct states} 
\label{sec:2}

Here we detail the Bethe ansatz results for the overlap of the Bethe states (cf.~\eref{ba-eig}) 
with the zero-momentum (one-site shift invariant) N\'eel state $|N\rangle$ (cf.~\eref{neel}) 
and the Majumdar-Ghosh (MG) $|MG\rangle$ state (cf.~\eref{mg}). In particular, we specialize 
the Bethe ansatz results to the case of eigenstates described by perfect strings.

@@

We stard discussing the overlaps with the N\'eel state. Due to the zero-momentum 
constraint, only parity-invariant Bethe states can have non-zero 
N\'eel overlap~\cite{brockmann-2014,wouters-2014}. The corresponding solutions of the Bethe 
equations~\eref{ba-eq} contain only pairs of rapidities with opposite sign.  Here we denote the 
generic parity-invariant rapidity configuration as $|\{\pm\tilde\lambda_j\}_{j=1}^m,n_\infty
\rangle$, i.e., considering only positive rapidities (as stressed by the tilde in $\tilde
\lambda_j$). Here $m$ is the number of rapidity pairs. Since the N\'eel state is not invariant 
under $SU(2)$ rotations, eigenstates with infinite rapidities can have non-zero N\'eel overlaps.
We denote the number of infinite rapidities as $N_{\infty}$. Note that one has $M=L/2=
N_\infty+2m$. The density of infinite rapidities is denoted as $n_\infty\equiv N_\infty/L$. 
The overlap between the Bethe states and the Neel state $|N\rangle$ reads~\cite{brockmann-2014,
pozsgay-2014} 
%
\begin{equation}
\label{neel-ov}
\frac{\langle N|\{\pm\tilde\lambda_j\}_{j=1}^m,n_\infty\rangle}{|||\{\tilde\lambda_j\}_{j=1}^m,
n_\infty\rangle||}=\frac{\sqrt{2}N_{\infty}!}{\sqrt{(2N_\infty)!}}\left[\prod_{j=1}^m
\frac{\sqrt{\tilde\lambda_j^2+1}}{4\tilde\lambda_j}\right]\sqrt{\frac{\textrm{det}_m(G^+)}{
\textrm{det}_m(G^-)}}.
\end{equation}
%
The matrix $G^\pm$ is  defined as  
%
\begin{equation}
\label{G-pm}
\fl\quad G^{\pm}_{jk}=\delta_{jk}\Big(LK_{1/2}(\tilde\lambda_j)-\sum\limits_{l=1}^mK_1^+(
\tilde\lambda_j,\tilde\lambda_l)\Big)+K_{1}^{\pm}(\tilde\lambda_j,\tilde\lambda_k),
\quad\,j,k=1,\dots,m, 
\end{equation}
%
where 
%
\begin{equation}
\label{K}
K_1^\pm(\lambda,\mu)=K_1(\lambda-\mu)\pm K_1(\lambda+\mu) \quad\textrm{with}\quad 
K_\alpha(\lambda)\equiv\frac{8\alpha}{\lambda^2+4\alpha^2}. 
\end{equation}
%
Note that our definitions of $K_{\alpha}(\lambda)$ differs from the one in 
Ref.~\cite{brockmann-2014}, due to a factor $2$ in the definition of the 
rapidities (cf.~\eref{str-hyp}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The low-density regime ({\bf PROBLEM !!!})}
\label{sec:low-density}

It is interesting to consider the situation with $M/L\to\infty$ (low-density limit). 
We first consider the case with all the particles forming a single zero-momentum 
string. Due to parity invariance only strings of even length are allowed. 
We assume the string hypothesis~\eref{str-hyp}, restricting ourselves to zero-momentum 
strings, i.e., 
%
\begin{equation}
\lambda^{j}_{n}=-i(n-1-2j)+i\delta_{n}^j,
\end{equation}
%
with $n$ the length of the string and $j$ labelling the different string components. 
Due to parity invariance we restrict ourselves to half strings, i.e., $j=1,\dots,n/2$.
We verified by explicit computation that for a single $m$ string one has (at leading order 
in the string deviations) 
%
\begin{equation}
\det(G^+)=-\frac{2^{n/2}}{\delta_n^{n/2}}L\prod_{j>k=1}^{n/2-1}(\delta_n^k-\delta_n^j)^{-1},
\end{equation}
%
and 
%
\begin{equation}
\det(G^-)=-\frac{2^{n/2}}{\delta_n^{n/2}}(L-1)\prod_{j>k=1}^{n/2-1}(\delta_n^k-\delta_n^j)^{-1}.
\end{equation}
%
This implies that 
%
\begin{equation}
\frac{\det_n(G^+)}{\det_n(G^-)}\to\frac{L}{L-1},\quad\forall n.
\end{equation}
%
However, it is straightforward to check that  
%
\begin{equation}
\prod_{j=1}^{n/2}\frac{\sqrt{\lambda_j^2+1}}{4\lambda_j}\propto\sqrt{\delta_n^{n/2}}.
\end{equation}
%
{\bf The prefactor does not seem to be a simple function of $n$}.

We conclude that Bethe states corresponding to a single string have zero N\'eel 
overlap.

A problem arises if one considers the Majumdar-Ghosh state. The overlap is obtained 
rescaling the N\'eel overlap (see below) with the prefactor
%
\begin{equation}
\prod\limits_{j=1}^n\frac{1}{2}
\Big(1-\frac{\lambda_j-i}{\lambda_j+i}\Big). 
\end{equation}
%
Note that the product above is over all the string components. 
For a single string one gets 
%
\begin{equation}
\prod\limits_{j=1}^n\frac{1}{2}
\Big(1-\frac{\lambda_j-i}{\lambda_j+i}\Big)\propto\frac{1}{\delta_n^{n/2}}.
\end{equation}
%
Putting everything together one would get a diverging MG overlap which is clearly wrong.

@@

%%%QA FULL DYNAMICS
\bibitem{de-nardis-2015}
J.~De~Nardis, L.~Piroli, and J.-S.~Caux, arXiv:1505.03080.

\bibitem{de-nardis-2015a}
J.~De~Nardis and J.-S.~Caux, J.\ Stat.\ Mech.\ (2014) P12012.

%%% ETH
\bibitem{deutsch-1991} 
J.~M.~Deutsch, Phys.\ Rev.\ A\ {\bf 43}, 2046 (1991).

\bibitem{srednicki-1994}
M.~Srednicki, Phys.\ Rev.\ E\ {\bf 50}, 888 (1994).
